Interactive Visualization of Hierarchically Structured Data

### 2016-12-05 treelapse.Rmd

Now that we have specified the essential questions of interest, we survey some
ideas from the visualization literature that can be applied to answer
them. Again, the core difficulty is one of high dimensionality, so it should be
no surprise that the techniques we adapt come from the literature on
high-dimensional data visualization, namely, work on the focus-plus-context and
linking.

The focus-plus-context principle is that it is possible to study large collections
of visual elements at multiple scales, and these scales can be juxtaposed by
"focusing" on a few elements of interest while maintaining the "context" of the
coarser-scale background. A simple example of this idea is to include a search
box that highlights samples corresponding to matching terms (the focus) while
muting those that don't (the context). Two more sophisticated methods anchored
in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are central
to the proposals in treelapse \@cite(hochheiser2004dynamic,
heer2004doitrees). In timeboxes, a collection of time series are graphically
queried using interactive brushes. Series that pass through all of the
user-specified brushes are highlighted, and the rest are faded into the
background. Hence, time series meeting the constraints imposed by the brushes
are focused, while the remainder remain present, though de-emphasized, as
context. This method can be interpreted programmatically as the visual version
of a database query, or probabilistically as the conditional distribution for
the full series, given it passes through certain bounds.

In DOI Trees, the viewer's attention is focused on a collection of
high-interest nodes, while the remaining lower-interest nodes are left on the
fringes (or entirely hidden) as context. The implementation is modularized into
two tasks -- the determination of a DOI distribution over nodes in the tree and
visual layout of a tree given DOI assignments. The DOI distribution used in
\@cite(heer2004doitrees) places maximal interest on nodes that the user clicks,
along with all ancestors. The DOI for all other nodes is defined as the
distance to the closest maximal interest node. The layout proposed here trimmed
low-interest subtrees until the remaining nodes all fit within the screen
size. By changing the minimal DOI below which nodes are hidden, the user can
transition between node-specific and full-tree scales.

In linking, alternative representations of the same samples are placed side by
side in a way that suggests covariation across the representations. A canonical
application is to linked scatterplot brushing \@cite(becker1987brushing). Here,
a scatterplot matrix is constructed displaying the relationship between all
pairs of variables.  Points brushed in one scatterplot are then highlighted in
all others. For example, this helps the user determine whether an outlier in
one dimension is an outlier in others. Another instance of this idea is in
linking the results of dimensionality reduction methods to displays of the raw
data, as implemented by XGobi and Cranvas, for example. As in timeboxes,
linking ca ben interpreted as database queries or conditional probabilities:
given a subset of the series after conditioning on the values for one set of
features, what are the values for a second set \@cite(buja1996interactive)?

Finally, while not exactly a fundamental visualization principle, we note that
our work is deliberately grounded in the R software ecosystem. Not only does
this make these methods more broadly accessible, we hope to encourage a more
animated exchange between data preparation and modeling with interactive
visualization. Moreover, our tools are intentionally limited in scope --
designed to facilitate this dialog for a specific class of problems, rather
than providing a toolbox for generic types of visualization design. We believe
that this narrow context within a broad ecosystem strikes a balance between
problem-specificity and ease-of-use.

@@

Our first proposed visualization technique is a minor modification of the DOI
Tree idea. The standard DOI Tree definition does not have any notion of data
defined at nodes, it is only used a device for navigating tree structures. A
trivial extension can encode scalar data at nodes: make the node radius reflect
the associated scalar value. To reinforce this effect, we also encode this
scalar value in the width of the parent edge. When parent nodes have values
equal to the sum of their children, this creates the effect of values "flowing"
from the root to leaves. To help viewers make use of their domain knowledge, we
have included a search box that highlights paths to matching node terms. Edges
are ordered from widest on the left to narrowest on the right. While this
method can only be used to visualize a single scalar-value per tree node, it
suggests an approach to the tree-structured differential abundance problem,
which we call the DOI Sankey.

In the DOI Sankey, we split each edge in the DOI Tree across different
groups. For example, suppose we have the average counts for treatment and
control groups at each tree node. Every edge in the tree is split into two
colors^[We use the colorbrewer palette to facilitate readability], with the
relative widths of the different colors reflecting the difference in sizes for
the two groups. The overall width of each edge is represents the sum of values
across all groups.

This display is designed to facilitate investigation of the tree structured
differential abundance question. For example, for a single node and a single
group, we can compute the average abundance at that node among all samples in
that group. This will give the width for that group's color on the edge leading
to the specified node. Differentially abundant subtrees then correspond to
subtrees where some colors occupy more space than others. That is, this
representation makes it easier to identify points where the "flows" for
different groups diverge -- the colors begin to separate. The DOI principle
assists navigation across the tree structure, allowing focus on individual flow
structures without losing broader tree context.

Our third display is directed at the microbial dynamics question instead. It
ignores the DOI principle, placing emphasis rather on linking and the
representation of full time series. Here, two panels are arranged one over the
other; one displays all time series, while the other displays all tree nodes,
with node sizes reflecting the average time series value at that node. For this
reason, we call the display, Timebox Trees. In the time series panel, we have
directly implemented the timeboxes idea. Our link is between the two panels:
when a set of series is highlighted by the timeboxes, the associated tree nodes
are also highlighted. For example, timeboxes can be used to focus on a set of
series with specific shape -- increased abundance after an ecological shock,
for example -- and then identify along what subtrees this pattern is
present. To further focus attention, a pan-zoom scented widget is provided in
the top right [@willett2007scented]. The widget is a miniature version of
the full time series panel, equipped with a single brush whose extent specifies
the limits in the main time series panel. As in the DOI trees and Sankeys, a
search bar can be used to highlight of interest a priori.

The final display currently implemented in the package is the natural converse
of the Timebox Trees display. Rather than defining visual queries in terms of
time series, we define queries using nodes in the tree. For this reason, we
call the display Treeboxes. Rather than focusing on the intersection of
brushes, as in Timebox Trees, we focus on the union of brushed over nodes. This
allows us to highlight the series associated with nodes on distant
subtrees. This display is also suited for the microbial dynamics problem. For
example, by highlighting all nodes at one taxonomic level in the tree, we can
easily summarize the time series pattern for all the taxa at that
level. Alternatively, focusing on all the children below a single node makes it
possible to see how much correlation and competition there is between
taxonomically similar microbes. As in the Timebox Trees display, a search box
and pan-zoom scented widget is provided.

@@

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted its CSTs using
a heatmap, with microbes ordered according to a hierarchical
clustering.  By using the DOI Sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual microbes loses power -- we can compare
microbe abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

@@

Finally, we can study the behavior of the regression tree itself using
the DOI Sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, in Figure ?? focuses on 2011 vs. 2012
split among weekday samples, showing that this split distinguishes
between samples falling in the second and third quintiles of 8AM
demand.

>>>>

Now that we have specified the essential questions of interest, we
survey some ideas from the visualization literature that can be
applied to answer them. As the core difficulty is high-dimensionality,
so it should be no surprise that the techniques we adapt come from the
literature on high-dimensional data visualization, namely,
focus-plus-context and linking.

The focus-plus-context principle is that large collections of visual
elements can be studied at multiple scales, by simultaneously
focusing" on a few elements of interest and maintaining the "context"
of the coarser-scale background. A simple example of this idea is to
include a search box that highlights matching samples (focus) and
mutes the rest (context). Two more sophisticated methods anchored in
this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
central to the proposals in treelapse [@hochheiser2004dynamic,
heer2004doitrees]. In timeboxes, a collection of time series are
graphically queried using interactive brushes. Series that pass
through all of the user-specified brushes are highlighted, and the
rest are faded to the background. Hence, time series meeting the
constraints imposed by the brushes are focused, while the remainder
are de-emphasized, though they remain present as context. This method
can be interpreted programmatically as the visual analog of a database
query, or probabilistically as the conditional distribution for the
full series, given it passes through certain bounds.

In DOI trees, the viewer's attention is focused on a collection of
high-interest nodes, while the remaining lower-interest nodes are left
on the fringes as context. The implementation is modularized into two
tasks -- the determination of a DOI distribution over nodes in the
tree and visual layout of a tree given DOI assignments. The DOI
distribution used in [@heer2004doitrees] places maximal interest on
the node that the user had most recently clicked, along with all
ancestors. The DOI for all other nodes is defined as the distance to
the closest maximal interest node. The layout step then trims
low-interest subtrees until the remaining nodes fit within a given
screen size. By adjusting the minimal DOI below which nodes are
hidden, the user can transition between node-specific and full-tree
scales.

In linking, alternative representations of the same samples are placed
side-by-side in order to display covariation across views. A canonical
application is to linked scatterplot brushing
[@becker1987brushing]. Here, a scatterplot matrix gives the
relationship between all pairs of variables.  Points brushed in one
scatterplot are then highlighted in all others. For example, this
helps the user determine whether an outlier in one dimension is an
outlier in others. Another instance of this idea links the results of
dimensionality reduction methods to displays of the raw data, as
implemented by XGobi and Cranvas, for example [@xie2013cranvas,
@swayne1998xgobi]. As in timeboxes, linking can be interpreted as
database queries or conditional probabilities: given a subset of the
series after conditioning on the values for one set of features, what
are the values for a second set [@buja1996interactive]?

Finally, unrelated to established visualization principles, we note
that our work is deliberately grounded in the R software
ecosystem. Not only does this make these methods more broadly
accessible, we hope to facilitate exchange between data modeling and
interactive visualization. Moreover, our tools are intentionally
limited in scope -- designed to facilitate this dialog for a specific
class of problems, rather than providing a toolbox for generic types
of visualization design. We believe that this narrow context within a
broad ecosystem strikes a balance between problem-specificity and
ease-of-use.

@@

Our first proposed visualization technique is a minor modification of
the DOI tree. The standard DOI tree definition does not have any
notion of data defined at nodes, it is only used a device for
navigating tree structures. A trivial extension can encode scalar data
at nodes: have the node radius reflect the associated scalar value. To
reinforce this effect, we can adjust the width of the parent
edge. When parent nodes have values equal to the sum of their
children, this creates the effect of values "flowing" from the root to
leaves. To help viewers make use of their domain knowledge, we have
included a search box that highlights paths to nodes with matching
terms. Edges are ordered from widest on the left to narrowest on the
right. While this method can only represent a single scalar-value per
tree node, it suggests an approach to the tree-structured differential
abundance problem, which we call the DOI sankey.

In the DOI sankey, we split each edge in the DOI Tree across different
groups. For example, suppose we have the average counts for treatment
and control groups at each tree node. Every edge in the tree is split
into two colors^[We use the colorbrewer palette to facilitate
readability], with relative widths of the different colors reflecting
differences in sizes for the two groups. The overall width of each
edge represents the sum of values across all groups.

This display is designed to facilitate investigation of the tree
structured differential abundance question. For example, for a single
node and a single group, first compute the average abundance at that
node among all samples in that group. This will give the width for
that group's color on the edge leading to the specified
node. Differentially abundant subtrees then correspond to subtrees
where some colors occupy more space than others. That is, this
representation makes it easier to identify points where the "flows"
for different groups diverge -- the colors begin to separate. The DOI
principle assists navigation across the tree structure, allowing focus
on individual flow structures without losing broader tree context.

Our third display is directed at the microbial dynamics
question. Here, two panels are arranged one over the other; one
displays all time series, while the other displays all tree nodes,
with node sizes reflecting the time series value at that node. For
this reason, we call the display, Timebox Trees. In the time series
panel, we have directly implemented the timeboxes idea. We then link
the panels: when a set of series is highlighted by the timeboxes, the
associated tree nodes are also highlighted. For example, timeboxes can
be used to focus on a set of series with specific shape -- increased
abundance after an ecological shock, for example -- and identify along
what subtrees this pattern is present. To further focus on specific
elements, a pan-zoom scented widget is provided
[@willett2007scented]. The widget is a miniature version of the full
time series panel, equipped with a single brush whose extent specifies
the limits in the main time series panel. As in the DOI trees and
sankeys, a search bar can be used to highlight those series of
interest a priori.

The final display currently implemented in the package is the natural
converse of the Timebox Trees display. Rather than defining visual
queries in terms of time series, it defines queries using nodes in the
tree. For this reason, we call the display Treeboxes. Rather than
focusing on the intersection of brushes, as in Timebox Trees, we focus
on the union of brushed over nodes. This allows us to highlight series
associated with nodes on distant subtrees. This display is also suited
for the microbial dynamics problem. For example, by highlighting all
nodes at one taxonomic level in the tree, we can easily summarize the
time series pattern for all the taxa at that level. Alternatively,
focusing on all the children below a single node makes it possible to
see how much correlation and competition there is between
taxonomically similar microbes. As in the Timebox Trees display, a
search box and pan-zoom scented widget are provided.

@@

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted its CSTs using
a heatmap, with microbes ordered according to a hierarchical
clustering.  By using the DOI sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual microbes loses power -- we can compare
microbe abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

@@

Finally, we can study the behavior of the regression tree itself using
the DOI sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, in Figure ?? focuses on 2011 vs. 2012
split among weekday samples, showing that this split distinguishes
between samples falling in the second and third quintiles of 8AM
demand.
